Responsive GPU Workgroup Management in the Presence of Persistent Kernels

### 2016-11-08 write-ups/PLDI/paper.tex
Got a bit further with intro.

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of threads.  In the context of irregular data
parallel algorithms accelerated on a GPU, this translates to requiring
fair scheduling of workgroups.  Without fairness, a workgroup may fail
to make progress, leading to starvation.  For example, if workgroup B
holds a mutex and workgroup A is spinning, waiting for the mutex to be
released, an unfair scheduler can forever schedule A, so that B never
releases a mutex.  Similarly, an unfair scheduler can induce
starvation if workgroup A is at an inter-workgroup barrier, spinning
until other workgroups arrive.

Current GPU programming models---CUDA~\cite{...}, OpenCL~\cite{...}
and HSA~\cite{...}---specify very few constraints on the manner in
which workgroups are scheduled.  In particular, they do \emph{not}
guarantee fair scheduling, and fair scheduling is not observed in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.  If a kernel is executed
with more workgroups than there are compute units then at least two
workgroups must share a single compute unit.  The simplest way to
handle this is an \emph{occupancy-bound} execution model, whereby once
a workgroup has commenced execution on a compute unit (is has become
\emph{occupant}), the workgroup has exclusive access to the compute
unit until it has finished execution, after which a subsequent
workgroup can be bound to the compute unit.  There is empirical
evidence that the occupancy-bound execution model is widely
implemented in practice~\cite{...}.

Clearly, the occupancy-bound execution model does not provide fair
scheduling between workgroups in general: if all compute units are
occupied then a not-yet-occupant workgroup will not be scheduled until
some occupant workgroup has completed execution.  Yet the execution
model \emph{does} provide fair scheduling between occupant workgroups,
which are bound to separate compute units that operate in parallel.
Current implementations of irregular data parallel algorithms on GPUs
assume this occupancy-bound execution model, exploiting it by
scheduling no more workgroups than there are available compute units.
Because current GPUs do provide the occupancy-bound execution model in
practice, this strategy allows starvation-free blocking
synchronization between workgroups.

However, none of the current GPU programming models mandate the
occupancy-bound execution model, and there are clear reasons why this
model is not actually desirable from the point-of-view of GPU vendors
nor end users.  First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all workgroups, the GPU cannot be used for other
tasks, such as graphics rendering or shorter, higher-priority
computations, in the meantime.  Precisely because of this problem,
operating systems employ a GPU \emph{watchdog} to heuristically detect
and kill long-running computations on the GPU that is being used for
rendering.  If one disables this watchdog and launches a long-running
kernel, the display becomes unresponsive for the duration of kernel
execution.  Second, \emph{energy throttling} is an important concern
for energy-constrained devices, such as smartphones and tablets.  In
the context of mobile GPU computing, it is desirable for a GPU driver
to be able to power down one or more compute units if the battery
level is low or if the user puts the device into a low power state.
Implementing energy throttling requires being able to suspend the
execution of a workgroup, power down the compute unit to which the
workgroup was bound, and schedule the workgroup for later execution on
one of the other compute units when it becomes free.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts~\cite{CitePersonalCommunication}, is that GPU
vendors are reluctant to commit to the occupancy-bound execution model
that they currently implement, due to the need for multi-tasking and
energy throttling, yet do not wish to guarantee fair scheduling, due
to the runtime overhead associated with supporting full preemption of
workgroups, e.g.\ based on time-slicing.  Vendors instead wish to
retain the essence of the occupancy-bound execution model, which is
simple to implement efficiently, providing preemption for the special
cases associated with multi-tasking for high-priority jobs (such as
graphics) and energy throttling.  In particular, they do not want to
commit to fair scheduling of a workgroup once it has been preempted.
As argued above, without fairness it is not possible to implement
non-blocking algorithms, including data parallel irregular algorithms,
safely.

@@

  Non-Functional: The cooperative kernel must ensure that offer_kill
  is executed frequently enough such that the system can accomodate
  soft-real time constraints (e.g. allowing a smooth framerate for
  graphics or throttling the energy before the mobile battery drains).
  To enable the cooperative execute efficiently, the scheduler should
  adopt a policy that avoids under utililasion of hardware resources
  for the cooperative kernel (e.g. the scheduler should not kill
  workgroups unless they are required).

>>>>

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of threads.  In the context of irregular data
parallel algorithms accelerated on a GPU, this translates to requiring
fair scheduling of workgroups.  Without fairness, a workgroup may fail
to make progress, leading to starvation.  For example, if workgroup A
holds a mutex and workgroup B is spin-waiting for the mutex to be
released, an unfair scheduler can forever schedule B, so that A never
releases the mutex.  Similarly, an unfair scheduler can induce
starvation if workgroup A is at an inter-workgroup barrier awaiting
the arrival of other workgroups, by allowing A to spin forever so that
the other workgroups do not progress to the barrier.

Current GPU programming models---CUDA~\cite{...}, OpenCL~\cite{...}
and HSA~\cite{...}---specify very few constraints on the manner in
which workgroups are scheduled.  In particular, they do \emph{not}
guarantee fair scheduling, and fair scheduling is not observed in
practice.  Roughly speaking, each workgroup executing a GPU kernel is
mapped to a hardware \emph{compute unit}.  If a kernel is executed
with more workgroups than there are compute units then at least two
workgroups must share a single compute unit.  The simplest way to
handle this is an \emph{occupancy-bound} execution model, whereby once
a workgroup has commenced execution on a compute unit (is has become
\emph{occupant}), the workgroup has exclusive access to the compute
unit until it has finished execution, after which a subsequent
workgroup can be bound to the compute unit.  There is empirical
evidence that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{...}, and it is
evidently a simple model to implement efficiently.

Clearly, the occupancy-bound execution model does not provide fair
scheduling between workgroups in general: if all compute units are
occupied then a not-yet-occupant workgroup will not be scheduled until
some occupant workgroup completes execution.  Yet the execution
model \emph{does} provide fair scheduling between \emph{occupant} workgroups,
because they are bound to separate compute units that operate in parallel.
Current implementations of irregular data parallel algorithms on GPUs
assume this occupancy-bound execution model, exploiting it by
scheduling no more workgroups than there are available compute units.
Because current GPUs do provide the occupancy-bound execution model in
practice, this strategy allows starvation-free blocking
synchronization between workgroups.

However, none of the current GPU programming models actually mandate
the occupancy-bound execution model, and there are clear reasons why
this model is not actually desirable from the point-of-view of GPU
vendors nor end users.  First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all workgroups, the GPU cannot be used for other
tasks, such as graphics rendering or shorter, higher-priority
computations, in the meantime.  Precisely because of this problem,
operating systems employ a GPU \emph{watchdog} to heuristically detect
and kill long-running computations on the GPU that is being used for
rendering.  If one disables this watchdog and launches a long-running
kernel, the display becomes unresponsive for the duration of kernel
execution.  Second, \emph{energy throttling} is an important concern
for battery-powered devices, such as smartphones and tablets.  It is
desirable for a mobile GPU driver to be able to power down one or more
compute units if the battery level is low or if the user puts the
device into a low power state.  Implementing energy throttling
requires being able to suspend the execution of a workgroup, power
down the compute unit to which the workgroup was bound, and schedule
the workgroup for later execution on a different compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts~\cite{CitePersonalCommunication}, is that GPU
vendors are reluctant to commit to the occupancy-bound execution model
that they currently implement, due to the need for multi-tasking and
energy throttling, yet do not wish to guarantee fair scheduling, due
to the runtime overhead associated with supporting full preemption of
workgroups, e.g.\ based on time-slicing.  Vendors instead wish to
retain the essence of the occupancy-bound execution model, which is
simple to implement efficiently, but provide preemption for the special
cases associated with multi-tasking for high-priority jobs (most crucially
graphics) and energy throttling.  In particular, they do not want to
commit to fair scheduling of a workgroup once it has been preempted.
As argued above, without fairness it is not possible to implement
non-blocking algorithms, including data parallel irregular algorithms,
safely.

\paragraph{Our proposal}

@@

  Non-Functional: The cooperative kernel must ensure that offer\_kill
  is executed frequently enough such that the system can accomodate
  soft-real time constraints (e.g. allowing a smooth framerate for
  graphics or throttling the energy before the mobile battery drains).
  To enable the cooperative execute efficiently, the scheduler should
  adopt a policy that avoids under utililasion of hardware resources
  for the cooperative kernel (e.g. the scheduler should not kill
  workgroups unless they are required).
